---
title: "Report Italian Vaccines"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r include = FALSE, message=FALSE, warning = FALSE}
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(tseries)) install.packages("tseries", repos = "http://cran.us.r-project.org")
if(!require(gtools)) install.packages("gtools", repos = "http://cran.us.r-project.org")
if(!require(sqldf)) install.packages("sqldf", repos = "http://cran.us.r-project.org")

library(dplyr)
library(gtools)
library(tidyverse)
library(ggplot2)
library(tseries)
library(sqldf)

```

## 1 Total non-vaccinated Population



```{r include = FALSE, message=FALSE, warning = FALSE}

url <- "https://github.com/AyrTraveler/Italian_Vaccination_Program/raw/main/italian_vaccination.csv"
dat <- read_csv(url) %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = first_dose, males = sum(males), females = sum(females))

dat  <- dat %>% filter(administration_date > "2020-12-31")


cumulativeSum = function(x){
  
  cumsum(ifelse(is.na(x),0,x))
  
}


clean_set  =  dat %>%  group_by(administration_date, age_range) %>%
  summarize(dose = sum(dose))
clean_set <- clean_set %>% spread(age_range,dose)
ADMIN_DATE <- as.data.frame(clean_set) %>% select(administration_date)
clean_set <- as.data.frame(clean_set) %>% select(-administration_date)
clean_set <- apply(as.matrix(clean_set),2,cumulativeSum)
clean_set <- as.data.frame(clean_set)  %>% gather(age_range, dose, `12-19`:`90+`)
clean_set <- data.frame(ADMIN_DATE,clean_set)

first_set <- clean_set %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>%
  summarise(dose = sum(dose)) %>%
  mutate(population = 60.0)


final_set <- first_set
final_set <- final_set %>% 
  mutate(difference = (population-dose)/population*100) %>% 
  mutate(age_range = "") %>% 
  mutate(Type = "First doses")

dat <- read_csv(url) %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = second_dose, males = sum(males), females = sum(females))

dat  <- dat %>% filter(administration_date > "2020-12-31")


cumulativeSum = function(x){
  
  cumsum(ifelse(is.na(x),0,x))
  
}


clean_set  =  dat %>%  group_by(administration_date, age_range) %>%
  summarize(dose = sum(dose))
clean_set <- clean_set %>% spread(age_range,dose)
ADMIN_DATE <- as.data.frame(clean_set) %>% select(administration_date)
clean_set <- as.data.frame(clean_set) %>% select(-administration_date)
clean_set <- apply(as.matrix(clean_set),2,cumulativeSum)
clean_set <- as.data.frame(clean_set)  %>% gather(age_range, dose, `12-19`:`90+`)
clean_set <- data.frame(ADMIN_DATE,clean_set)

first_set <- clean_set %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>%
  summarise(dose = sum(dose)) %>%
  mutate(population = 60.0)


temp_set <- first_set
temp_set <- temp_set %>% 
  mutate(difference = (population-dose)/population*100) %>% 
  mutate(age_range = "") %>% 
  mutate(Type = "Second doses")

final_set <- bind_rows(final_set, temp_set)

```


```{r echo= FALSE, message=FALSE, warning = FALSE}
final_set %>% 
  ggplot(aes(reorder(Type,difference) ,difference, label = round(population-dose,2))) +
  geom_bar(stat="identity", position = position_dodge(), show.legend = TRUE, fill = "darkcyan", width = 0.3) +
  geom_label(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(y= "% Population that has not yet received the first/second dose", x = "") 


```



## 2 Non-vaccinated population by age range

### 2.1 Non-vaccinated population by age range - First Doses


```{r include = FALSE, message=FALSE, warning = FALSE}

url <- "https://github.com/AyrTraveler/Italian_Vaccination_Program/raw/main/italian_vaccination.csv"
dat <- read_csv(url) %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = first_dose, males = sum(males), females = sum(females))

dat  <- dat %>% filter(administration_date > "2020-12-31")


cumulativeSum = function(x){
  
  cumsum(ifelse(is.na(x),0,x))
  
}


clean_set  =  dat %>%  group_by(administration_date, age_range) %>%
  summarize(dose = sum(dose))
clean_set <- clean_set %>% spread(age_range,dose)
ADMIN_DATE <- as.data.frame(clean_set) %>% select(administration_date)
clean_set <- as.data.frame(clean_set) %>% select(-administration_date)
clean_set <- apply(as.matrix(clean_set),2,cumulativeSum)
clean_set <- as.data.frame(clean_set)  %>% gather(age_range, dose, `12-19`:`90+`)
clean_set <- data.frame(ADMIN_DATE,clean_set)

first_set <- clean_set %>% 
  filter(!(age_range %in% c('70-79','80-89','90+'))) %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>%
  mutate( population =
            case_when(
              
              age_range == "12-19" ~ 4.635324714,
              age_range == "20-29" ~ 6.720621333,
              age_range == "30-39" ~ 7.1803445,
              age_range == "40-49" ~ 8.792986,
              age_range == "50-59" ~ 9.113997,
              age_range == "60-69" ~ 7.6751905
              
            ))

second_set <- clean_set %>% 
  filter(age_range %in% c('70-79','80-89','90+')) %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>% 
  mutate(age_range= ">70") %>% 
  group_by(age_range) %>% 
  summarize(dose = sum(dose)) %>% 
  mutate(population = 10.5243685)
  

final_set <- bind_rows(first_set,second_set)
final_set <- final_set %>% mutate(difference = (population-dose)/population*100)

third_set <- clean_set %>% 
  filter(!(age_range %in% c('70-79','80-89','90+'))) %>% 
    mutate( population =
            case_when(
              
              age_range == "12-19" ~ 4.635324714,
              age_range == "20-29" ~ 6.720621333,
              age_range == "30-39" ~ 7.1803445,
              age_range == "40-49" ~ 8.792986,
              age_range == "50-59" ~ 9.113997,
              age_range == "60-69" ~ 7.6751905
              
            ))

fourth_set <- clean_set %>% 
  filter(age_range %in% c('70-79','80-89','90+')) %>% 
  mutate(age_range= ">70") %>% 
  group_by(administration_date,age_range) %>% 
  summarise(dose = sum(dose)) %>% 
  mutate(population = 10.5243685)



trend_set <- bind_rows(third_set,fourth_set)
trend_set <- trend_set %>% mutate(difference = (population-dose/1000000)/population*100)

pivot <- trend_set %>% group_by(age_range) %>% summarise(min_diff = min(difference))
trend_set <- sqldf('
                   SELECT T.*,P.min_diff
                   FROM trend_set T
                   LEFT JOIN pivot P
                   ON T.age_range = P.age_range
                   AND T.difference = P.min_diff
                   
                   ')


```

The plot shows percentage/millions of people that have not yet received the first vaccine shot, by age range:

```{r echo= FALSE, message=FALSE, warning = FALSE}
final_set %>% 
  ggplot(aes(reorder(age_range,difference) ,difference, label = round(population-dose,2))) +
  geom_bar(stat="identity", position = position_dodge(), show.legend = TRUE, fill = "darkcyan", width = 0.7) +
  geom_label(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(y= "% Population that has not yet received the first dose", x = "Age Range") 


```

The plot shows the trend of the percentage of population who has not yet received the first vaccine shot, by age range:

```{r echo= FALSE, message=FALSE, warning = FALSE}

trend_set %>% 
  ggplot(aes(administration_date,difference, label = round(min_diff,1))) +
  geom_point(size = 0.2, color = "darkcyan") +
  geom_label(show.legend = FALSE, size = 3)+
  facet_wrap(~age_range) +
  scale_y_continuous(position = "right") +
  theme(legend.position = "left") +
  theme_minimal()+
  labs(y= "Remaining doses [%]", x = "") 


```


### 2.2 Non-vaccinated population by age range - Second Doses


```{r include = FALSE, message=FALSE, warning = FALSE}

url <- "https://github.com/AyrTraveler/Italian_Vaccination_Program/raw/main/italian_vaccination.csv"
dat <- read_csv(url) %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = second_dose, males = sum(males), females = sum(females))

dat  <- dat %>% filter(administration_date > "2020-12-31")


cumulativeSum = function(x){
  
  cumsum(ifelse(is.na(x),0,x))
  
}


clean_set  =  dat %>%  group_by(administration_date, age_range) %>%
  summarize(dose = sum(dose))
clean_set <- clean_set %>% spread(age_range,dose)
ADMIN_DATE <- as.data.frame(clean_set) %>% select(administration_date)
clean_set <- as.data.frame(clean_set) %>% select(-administration_date)
clean_set <- apply(as.matrix(clean_set),2,cumulativeSum)
clean_set <- as.data.frame(clean_set)  %>% gather(age_range, dose, `12-19`:`90+`)
clean_set <- data.frame(ADMIN_DATE,clean_set)

first_set <- clean_set %>% 
  filter(!(age_range %in% c('70-79','80-89','90+'))) %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>%
  mutate( population =
            case_when(
              
              age_range == "12-19" ~ 4.635324714,
              age_range == "20-29" ~ 6.720621333,
              age_range == "30-39" ~ 7.1803445,
              age_range == "40-49" ~ 8.792986,
              age_range == "50-59" ~ 9.113997,
              age_range == "60-69" ~ 7.6751905
              
            ))

second_set <- clean_set %>% 
  filter(age_range %in% c('70-79','80-89','90+')) %>% 
  group_by(age_range) %>% 
  summarise(dose = max(dose)/1000000) %>% 
  mutate(age_range= ">70") %>% 
  group_by(age_range) %>% 
  summarize(dose = sum(dose)) %>% 
  mutate(population = 10.5243685)
  

final_set <- bind_rows(first_set,second_set)
final_set <- final_set %>% mutate(difference = (population-dose)/population*100)

third_set <- clean_set %>% 
  filter(!(age_range %in% c('70-79','80-89','90+'))) %>% 
    mutate( population =
            case_when(
              
              age_range == "12-19" ~ 4.635324714,
              age_range == "20-29" ~ 6.720621333,
              age_range == "30-39" ~ 7.1803445,
              age_range == "40-49" ~ 8.792986,
              age_range == "50-59" ~ 9.113997,
              age_range == "60-69" ~ 7.6751905
              
            ))

fourth_set <- clean_set %>% 
  filter(age_range %in% c('70-79','80-89','90+')) %>% 
  mutate(age_range= ">70") %>% 
  group_by(administration_date,age_range) %>% 
  summarise(dose = sum(dose)) %>% 
  mutate(population = 10.5243685)



trend_set <- bind_rows(third_set,fourth_set)
trend_set <- trend_set %>% mutate(difference = (population-dose/1000000)/population*100)

pivot <- trend_set %>% group_by(age_range) %>% summarise(min_diff = min(difference))
trend_set <- sqldf('
                   SELECT T.*,P.min_diff
                   FROM trend_set T
                   LEFT JOIN pivot P
                   ON T.age_range = P.age_range
                   AND T.difference = P.min_diff
                   
                   ')



```

The plot shows percentage/millions of people that have not yet received the second vaccine shot, by age range:

```{r echo= FALSE, message=FALSE, warning = FALSE}
final_set %>% 
  ggplot(aes(reorder(age_range,difference) ,difference, label = round(population-dose,2))) +
  geom_bar(stat="identity", position = position_dodge(), show.legend = TRUE, fill = "darkcyan", width = 0.7) +
  geom_label(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(y= "% Population that has not yet received the first dose", x = "Age Range") 


```

The plot shows the trend of the percentage of population who has not yet received the second vaccine shot, by age range:

```{r echo= FALSE, message=FALSE, warning = FALSE}

trend_set %>% 
  ggplot(aes(administration_date,difference, label = round(min_diff,1))) +
  geom_point(size = 0.2, color = "darkcyan") +
  geom_label(show.legend = FALSE, size = 3)+
  facet_wrap(~age_range) +
  scale_y_continuous(position = "right") +
  theme(legend.position = "left") +
  theme_minimal()+
  labs(y= "Remaining doses [%]", x = "") 


```


## Non-vaccinated population by region

### 3.1 Non-vaccinated population by region - First Doses


```{r include = FALSE, message=FALSE, warning = FALSE}

url <- "https://github.com/AyrTraveler/Italian_Vaccination_Program/raw/main/italian_vaccination.csv"
dat <- read_csv(url)
first <- dat %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = first_dose, males = sum(males), females = sum(females)) %>% 
  filter(administration_date > "2020-12-31") %>% 
  group_by(region) %>% 
  summarize(dose = sum(dose)) %>% 
  mutate( population =
            case_when(
              
              region == "ABR" ~ 1.293941,
              region == "BAS" ~ 0.553254,
              region == "CAL" ~ 1.894110,
              region == "CAM" ~ 5.712143,
              region == "EMR" ~ 4.464119,
              region == "FVG" ~ 1.206216,
              region == "LAZ" ~ 5.755700,
              region == "LIG" ~ 1.524826,
              region == "LOM" ~ 10.027602,
              region == "MAR" ~ 1.512672,
              region == "MOL" ~ 0.300516,
              region == "PAB" ~ 0.533349,
              region == "PAT" ~ 0.545497,
              region == "PIE" ~ 4.311217,
              region == "PUG" ~ 3.953305,
              region == "SAR" ~ 1.611621,
              region == "SIC" ~ 4.875290,
              region == "TOS" ~ 3.692555,
              region == "UMB" ~ 0.870165,
              region == "VDA" ~ 0.125034,
              region == "VEN" ~ 4.879133,
              
            ),
          dose = dose/1000000,
          difference = (population-dose)/population*100
  ) %>%
   mutate(difference = round(difference,2))

day_span <- 7
second <- dat  %>%
  group_by(region) %>% 
  summarize(max_date =max(administration_date))  %>% 
  select(region,max_date)

third <- 
  dat %>% left_join(second, by = "region") %>% 
  filter(administration_date >= as.Date(max_date) - day_span & administration_date <= as.Date(max_date)) %>% 
  group_by(region) %>% 
  summarize(administration_date = max(administration_date), speed =sum(first_dose)/day_span/1e6 ) %>% 
  left_join(first, by = "region") %>% 
  mutate(final_date = as.Date(administration_date) + round((population-dose)/speed,0)) %>% 
  select(region,final_date)

third <- third[order(third$final_date),]


```

The plot shows percentage/millions of people that have not yet received the first vaccine shot, by region:

```{r echo= FALSE, message=FALSE, warning = FALSE}
first %>% 
  ggplot(aes(reorder(region,difference) ,difference, label = round(population-dose,2))) +
  geom_bar(stat="identity", position = position_dodge(), show.legend = TRUE, fill = "darkcyan", width = 0.7) +
  geom_label(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(y= "% Population that has not yet received the first dose", x = "Region") 


```

The final date is an ideal scenario that tells when the first dose vaccination will be completed, by region, assuming the most recent weekly average first dose vaccination rate to be constant:

```{r echo= FALSE, message=FALSE, warning = FALSE}
knitr::kable(third)

```


### 3.2 Non-vaccinated population by region - Second Doses


```{r include = FALSE, message=FALSE, warning = FALSE}

url <- "https://github.com/AyrTraveler/Italian_Vaccination_Program/raw/main/italian_vaccination.csv"
first <- read_csv(url) %>% 
  group_by(administration_date,region,supplier, region_name, age_range) %>% 
  summarise(dose = second_dose, males = sum(males), females = sum(females)) %>% 
  filter(administration_date > "2020-12-31") %>% 
  group_by(region) %>% 
  summarize(dose = sum(dose)) %>% 
  mutate( population =
            case_when(
              
              region == "ABR" ~ 1.293941,
              region == "BAS" ~ 0.553254,
              region == "CAL" ~ 1.894110,
              region == "CAM" ~ 5.712143,
              region == "EMR" ~ 4.464119,
              region == "FVG" ~ 1.206216,
              region == "LAZ" ~ 5.755700,
              region == "LIG" ~ 1.524826,
              region == "LOM" ~ 10.027602,
              region == "MAR" ~ 1.512672,
              region == "MOL" ~ 0.300516,
              region == "PAB" ~ 0.533349,
              region == "PAT" ~ 0.545497,
              region == "PIE" ~ 4.311217,
              region == "PUG" ~ 3.953305,
              region == "SAR" ~ 1.611621,
              region == "SIC" ~ 4.875290,
              region == "TOS" ~ 3.692555,
              region == "UMB" ~ 0.870165,
              region == "VDA" ~ 0.125034,
              region == "VEN" ~ 4.879133,
              
            ),
          dose = dose/1000000,
          difference = (population-dose)/population*100
  ) %>%
   mutate(difference = round(difference,2))


day_span <- 7
second <- dat  %>%
  group_by(region) %>% 
  summarize(max_date =max(administration_date))  %>% 
  select(region,max_date)

third <- 
  dat %>% left_join(second, by = "region") %>% 
  filter(administration_date >= as.Date(max_date) - day_span & administration_date <= as.Date(max_date)) %>% 
  group_by(region) %>% 
  summarize(administration_date = max(administration_date), speed =sum(second_dose)/day_span/1e6 ) %>% 
  left_join(first, by = "region") %>% 
  mutate(final_date = as.Date(administration_date) + round((population-dose)/speed,0)) %>% 
  select(region,final_date)

third <- third[order(third$final_date),]


```


The plot shows percentage/millions of people that have not yet received the second vaccine shot, by region:

```{r echo= FALSE, message=FALSE, warning = FALSE}
first %>% 
  ggplot(aes(reorder(region,difference) ,difference, label = round(population-dose,2))) +
  geom_bar(stat="identity", position = position_dodge(), show.legend = TRUE, fill = "darkcyan", width = 0.7) +
  geom_label(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(y= "% Population that has not yet received the first dose", x = "Region") 


```

The final date is an ideal scenario that tells when the second dose vaccination will be completed, by region, assuming the most recent weekly average second dose vaccination rate to be constant:

```{r echo= FALSE, message=FALSE, warning = FALSE}
knitr::kable(third)

```
